On the use of `boilerplate' statistical analysis sections?

### 2021-05-13 manuscript/draft_manuscript.tex
update manuscript methods results

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={How many researchers use `boilerplate' statistical analysis sections?},
  pdfauthor={Nicole White, Thiru Balasubramaniam, Richi Nayak, Adrian Barnett},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering

\title{How many researchers use `boilerplate' statistical analysis sections?}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{An observational study of papers published in \emph{PLOS ONE}.}
\author{Nicole White, Thiru Balasubramaniam, Richi Nayak, Adrian Barnett}
\date{21/05/2020}

@@

An ideal statistical analysis will use appropriate methods to create
insights from the data and inform the research questions. Unfortunately
many current statistical analyses are far from ideal, with many
researchers using the wrong methods, misinterpreting the results, or
failing to adequately check their assumptions (Goodman
\protect\hyperlink{ref-Goodman2008}{2008}; Leek et al.
\protect\hyperlink{ref-Leek2017}{2017}). Some researchers take a
``mechanistic'' approach to statistics, copying the few methods they
know regardless of their appropriateness, and then going through the
motions of the analysis (Stark and Saltelli
\protect\hyperlink{ref-Stark2018}{2018}).

Many researchers lack adequate training in research methods, and
statistics is something they do with trepidation and even ignorance
(Altman \protect\hyperlink{ref-Altman1994}{1994}; King et al.
\protect\hyperlink{ref-King2019}{2019}). However, using the wrong
statistical methods can cause real harm (Altman
\protect\hyperlink{ref-Altman1994}{1994}) and bad statistical practices
are being to used abet weak science (Stark and Saltelli
\protect\hyperlink{ref-Stark2018}{2018}). Statistical mistakes are a key
source of waste in research and partly explain the current
reproducibility crisis in science (Allison et al.
\protect\hyperlink{ref-Allison2016}{2016}). Even when the correct
methods are used, many researchers fail to describe them adequately,
making it difficult to reproduce the results (Ernst and Albers
\protect\hyperlink{ref-Ernst2017}{2017}; Zhou and Skidmore
\protect\hyperlink{ref-Zhou2018}{2018}).

The International Committee of Medical Journal Editors recommend that
researchers should: ``Describe statistical methods with enough detail to
enable a knowledgeable reader with access to the original data to judge
its appropriateness for the study and to verify the reported results''
(ICJME \protect\hyperlink{ref-ICMJE2019}{2019}). Although the general
lack of statistical understanding from both authors and reviewers means
this recommendation may not be checked. A recent survey of editors found
that only 23\% of health and medical journals used expert statistical
review for all articles (Hardwicke and Goodman
\protect\hyperlink{ref-Hardwicke2020}{2020}), which was little different
from a survey from 22 years ago (Goodman, Altman, and George
\protect\hyperlink{ref-Goodman1998}{1998}).

Two statisticians on this paper (AB and NW) have heard researchers admit
that they sometimes copy-and-paste their statistical methods sections
from other papers, regardless of whether they are appropriate. The aim
of this paper is to use text-mining methods to estimate the extent that
researchers are using cut-and-paste or `boilerplate' statistical methods
sections. Use of these methods sections indicates that little thought
has gone into the statistical analysis.

@@

\emph{PLOS ONE} is a large open access journal that publishes original
research across a wide range of scientific fields. Articles must be in
English. Article submissions are handled by an academic editor who
selects peer reviewers based on their self-nominated area(s) of
expertise. Submissions do not undergo formal statistical review.
Instead, reviewers are required to assess submissions against several
publication criteria, including whether: ``Experiments, statistics, and
other analyses are performed to a high technical standard and are
described in sufficient detail''. Authors are encouraged to follow
published reporting guidelines to ensure that chosen statistical methods
are appropriate for the study design, and adequate details are provided
to enable independent replication of results. Reporting guidelines are
available from the EQUATOR network which has developed guidelines for
study designs commonly used in health research, and these guidelines
were developed to tackle poor statistical application and reporting in
health and medical journals (Altman and Simera
\protect\hyperlink{ref-Altman2016}{2016}).

All \emph{PLOS ONE} articles are freely accessible via the PLOS
Application Programming Interface (API). This functionality enabled us
to conduct semi-automated searches of full-text articles and analyse
data on individual records, including text content and general
attributes such as publication date and field(s) of research. To find
papers with a statistical methods section we used targeted API searches
followed by article filtering based on section headings. The data were
downloaded on xx-xxxx-xxx.

\emph{Step 1}: Targeted API searches. API searches were completed using
the R package `rplos' (Chamberlain, Boettiger, and Ram
\protect\hyperlink{ref-rplos}{2020}). Search queries targeted the
presence of analysis-related terms anywhere in the article. Search terms
combined the words ``data'' or ``statistical'' with one of:
``analysis'', ``analyses'', ``method'', ``methodology'' or
``model(l)ing''. Search terms were intended to be broad whilst keeping
search results to a manageable number for full-text review (see Step 2).
By allowing terms to appear anywhere within the article, we accounted
for the possibility of relevant text being placed in different sections,
for example, in the \emph{Material and Methods} section versus
\emph{Results}. Search results were indexed by a unique Digital Object
Identifier (DOI). Attribute data collected per DOI included journal
volume, subject classification(s) and total article views since
publication.

@@

The ANZCTR was established in 2005 as part of a coordinated global
effort to improve research quality and transparency in clinical trials
reporting; observational studies can also be registered. All studies
registered on ANZCTR are publicly available and can be searched via an
online portal (\url{https://www.anzctr.org.au/}). Details required for
registration follow a standardised template (reference or supp file),
which covers participant eligibility, the intervention(s) being
evaluated, study design and outcomes. Researchers are asked to provide a
``brief description'' of the sample size calculations, statistical
methods and planned analyses, although this section is not compulsory
(anzctr \protect\hyperlink{ref-ANZCTR}{2019}). The information provided
must be in English. Studies are reviewed by ANZCTR staff for
completeness of key information, which does not include the completeness
of the statistical methods sections. Studies are not peer reviewed.

All studies available on ANZCTR were downloaded on 1 February 2020 in
XML format. We used all the text available in the ``Statistical
methods'' section. We also collated basic information about the study
including the study type (interventional or observational), submission
date, number of funders and target sample size. These variables were
chosen as we believed they might influence the completeness of the
statistical methods section, because we expected larger studies and
those with funding to be more complete, and we also were interested in
changes over time. Studies prior to 2013 were excluded as the
statistical methods section appeared to be introduced in 2013. Some
studies were first registered on the alternative trial database
\emph{clinicaltrials.gov} and then also posted to ANZCTR. We excluded
these studies because they almost all had no completed statistical
methods section as this requested by \emph{clinicaltrials.gov}.

@@

Text cleaning aimed to standardise notation and statistical terminology,
whilst minimising changes to article style and formatting. Full details
of text cleaning steps undertaken and corresponding R code are provided
in Supplementary File X.

Mathematical notation including Greek letters was converted from Unicode
characters to plain text. For example, the Unicode characters
corresponding to \(\theta\) (\textless U+03B8\textgreater) was replaced
with `theta'. Similarly, common symbols outside of Unicode blocks
including `\%' (percent) and `\textless{}' (`less-than') were converted
into plain text, using functions available in the `textclean' package
(Rinker \protect\hyperlink{ref-textclean}{2018}). General formatting was
removed, this included carriage returns, punctuation marks, in-text
references (e.g.~``{[}42{]}'') centred equations, and other non-ascii
characters. Text contained inside brackets was retained in the dataset
to maximise content for analysis, with brackets removed.

We compiled an extensive list of statistical terms to standardise
descriptions of statistical methods reported across both datasets. An
initial list was compiled by calculating individual word frequencies and
identifying relevant terms that appeared at least 100 times. Further
terms were sourced from index searches of statistics reference textbooks
{[}ref{]}. The final list is provided as Supplementary Material.
Possible variants including plurals (e.g.~`chi-squares') unhyphenated
(e.g `chi square') and combined (e.g.~`chisquare') terms were
transformed to singular, hyphenated form (e.g `chi-square'). Common
statistical tests were also hyphenated (e.g.~`hosmer lemeshow' to
`hosmer-lemeshow').

As a final step, common stop words including pronouns, contractions and
selected prepositions were removed. We retained selected stop words
that, if excluded, may have changed the context of statistical method(s)
being described, for example `between' and `against'.

@@

@@

The statistical methods section for the ANZCTR data was somtimes missing
and so we examined if there were particular studies where this section
was more likely to be missing. We used a logistic regression model
fitted using a Bayesian paradigm. A small number of sections were
labelled as ``Not applicable'', ``Nil'' or ``None'' and we changed these
to missing.

@@

\hypertarget{plos-one}{%
\subsection{PLOS ONE}\label{plos-one}}

\begin{figure}
\centering
\includegraphics{draft_manuscript_files/figure-latex/unnamed-chunk-3-1.pdf}
\caption{\label{fig:plos-n}Search results by PLOS ONE volume (1st row);
word count per statistical methods section included in analysis (n =
111,731; 2nd row); subject classifications assigned to full-text records
included in analysis (3rd row)}
\end{figure}

API searches returned 131847 unique records, of which 111731 (85\%)
included a statistical methods section based on our search criteria.
Search results varied by journal volume (Figure 1A). The total number of
API search results peaked at volumes 8 (n = 19045) and 9 (n = 19045),
corresponding to years 2013 and 2014. This trend aligned with the total
number of papers published in PLOS ONE over the same timeframe. The
percentage of records that included a statistical methods section by
volume based on our proposed matching criteria varied between 64\%
(volume 2) and 86\% (volume 9).

The median length of statistical methods sections after text cleaning
was 127 per DOI (IQR: 61 to 254) (Figure 1B).

Across all records included in analysis, all included Biology and life
sciences (n = 107584), Earth sciences (n = 7605) and/or Computer and
information sciences (n = 5190) within their top 3 subject
classifications (Figure 1C).

{[}Figure here or as supplementary - 5 topic solution{]}

Results of topic modelling assuming five clusters showed clear
separation between statistical software, hypothesis testing and linear
modelling (Figure). The largest cluster, representing 48\% of all
included studies (Topic 2, n = 53,491).

Two topics (26\% of study sample) were dominated by terms related
statistical software, notably Prism/GraphPad (Topic 3, n = 10,940) and
SPSS (Topic 5, n = 18,029). Remaining topics focussed on hypothesis
testing (Topic, n = 12,531) and analysis of variance/multiple
comparisons (Topic 4, n = 16,740)

Top matches (Topic 3)

\begin{itemize}
\tightlist
\item
  10.1371/journal.pone.0043519: All statistical analyses were performed
  using GraphPad Prism (GraphPad Software Inc., San Diego, CA).
\item
  10.1371/journal.pone.0045453: GraphPad Prism (Graphpad Software, San
  Diego, CA) was used for all analyses.
\item
  10.1371/journal.pone.0170640: All statistical analysis of the data was
  performed using GraphPad Prism software (GraphPad Prism Software Inc.,
  La Jolla, Ca).
\item
  10.1371/journal.pone.0036750: All statistical analysis was performed
  using Graphpad Prism software.
\item
  10.1371/journal.pone.0058148: All statistical analysis was performed
  using the Graphpad Prism software.
\end{itemize}

Top matches (Topic 5) * 10.1371/journal.pone.0120046: Statistical
analysis was performed using the SPSS version 18.0 (SPSS Inc., Chicago,
IL, USA). Mann-Whitney test was used for the analysis of continuous
variables. Chi-square test, Fisher's exact test or one-way ANOVA test
was employed for the analysis of categorical variables. P \textless{}
0.05 was considered statistically significant.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={On the use of `boilerplate' statistical analysis sections?},
  pdfauthor={Nicole White, Thiru Balasubramaniam, Richi Nayak, Adrian Barnett},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs}
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}

\title{On the use of `boilerplate' statistical analysis sections?}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{An observational study of papers published in \emph{PLOS ONE} and studies posted to a trial registry.}
\author{Nicole White, Thiru Balasubramaniam, Richi Nayak, Adrian Barnett}
\date{21/05/2020}

@@

An ideal statistical analysis will use appropriate methods to create insights from the data and inform the research questions. Unfortunately many current statistical analyses are far from ideal, with many researchers using the wrong methods, misinterpreting the results, or failing to adequately check their assumptions ({\textbf{???}}; Leek et al. \protect\hyperlink{ref-Leek2017}{2017}). Some researchers take a ``mechanistic'' approach to statistics, copying the few methods they know regardless of their appropriateness, and then going through the motions of the analysis (Stark and Saltelli \protect\hyperlink{ref-Stark2018}{2018}).

Many researchers lack adequate training in research methods, and statistics is something they do with trepidation and even ignorance (Altman \protect\hyperlink{ref-Altman1994}{1994}; King et al. \protect\hyperlink{ref-King2019}{2019}).
However, using the wrong statistical methods can cause real harm (Altman \protect\hyperlink{ref-Altman1994}{1994}; Brown, Kaiser, and Allison \protect\hyperlink{ref-Brown2018}{2018}) and bad statistical practices are being to used abet weak science (Stark and Saltelli \protect\hyperlink{ref-Stark2018}{2018}).
Statistical mistakes are a key source of waste in research and partly explain the current reproducibility crisis in science (Allison et al. \protect\hyperlink{ref-Allison2016}{2016}). Even when the correct methods are used, many researchers fail to describe them adequately, making it difficult to reproduce the results (Ernst and Albers \protect\hyperlink{ref-Ernst2017}{2017}; Zhou and Skidmore \protect\hyperlink{ref-Zhou2018}{2018}).
Poor statistical methods might not be caught by reviewers, as they may not be qualified to judge the statistics.
A recent survey of editors found that only 23\% of health and medical journals used expert statistical review for all articles (Hardwicke and Goodman \protect\hyperlink{ref-Hardwicke2020}{2020}), which was little different from a survey from 22 years ago (Goodman, Altman, and George \protect\hyperlink{ref-Goodman1998}{1998}).

There is guidance for researchers on how to write up their statistical methods and results.
The International Committee of Medical Journal Editors recommend that researchers should: ``Describe statistical methods with enough detail to enable a knowledgeable reader with access to the original data to judge its appropriateness for the study and to verify the reported results'' (ICJME \protect\hyperlink{ref-ICMJE2019}{2019}).
More detailed guideance is given by the SAMPL and EQUATOR guidelines (Lang and Altman \protect\hyperlink{ref-Lang2013}{2013}; Altman and Simera \protect\hyperlink{ref-Altman2016}{2016}) with the latter covering all apsects of the paper. Both of these guidelines were led by Doug Altman, who spoke often and for many years about the need for better statistical reporting.
The awareness and use of these guidelines could be improved. There were 256 Google Scholar citations to the SAMPL paper (as at 15 March 2021) which is a good citation statistic for most papers, but is low considering the millions of papers that use statistical analysis.

Two statisticians on this paper (AB and NW) have heard researchers admit that they have copied-and-pasted their statistical methods sections from other papers, regardless of whether they are appropriate.
The aim of this paper is to use text-mining methods to estimate the extent that researchers are using cut-and-paste or `boilerplate' statistical methods sections.
Boilerplate text is that ``which can be reused in new contexts or applications without significant changes to the original'' (Wikipedia \protect\hyperlink{ref-Wikipedia}{2021}).
Use of these methods sections indicates that little thought has gone into the statistical analysis.

@@

\emph{PLOS ONE} is a large open access journal that publishes original research across a wide range of scientific fields. Articles must be in English. Article submissions are handled by an academic editor who selects peer reviewers based on their self-nominated areas of expertise. Submissions do not undergo formal statistical review. Instead, reviewers are required to assess submissions against several publication criteria, including whether: ``Experiments, statistics, and other analyses are performed to a high technical standard and are described in sufficient detail'' (PLOS \protect\hyperlink{ref-PLOS}{2021}). All reviewers are asked the question: ``Has the statistical analysis been performed appropriately and rigorously?'', with the possible responses of ``Yes'', ``No'' and ``I don't know''.

Authors are encouraged to follow published reporting guidelines such as EQUATOR, to ensure that chosen statistical methods are appropriate for the study design, and adequate details are provided to enable independent replication of results.

All \emph{PLOS ONE} articles are freely accessible via the PLOS Application Programming Interface (API). This enabled us to conduct semi-automated searches of full-text articles and analyse data on individual records, including text content and general attributes such as publication date and field(s) of research. To find papers with a statistical methods section we used targeted API searches followed by article filtering based on section headings. The data were downloaded on 3 July 2020.

\emph{Step 1}: Targeted API searches. API searches were completed using the R package `rplos' (Chamberlain, Boettiger, and Ram \protect\hyperlink{ref-rplos}{2020}). Search queries targeted the presence of analysis-related terms anywhere in the article. Search terms combined the words ``data'' or ``statistical'' with one of: ``analysis'', ``analyses'', ``method'', ``methodology'' or ``model(l)ing''. Search terms were intended to be broad whilst keeping search results to a manageable number for full-text review (see Step 2). By allowing terms to appear anywhere in the article, we accounted for the possibility of relevant text being placed in different sections, for example, in the \emph{Material and Methods} section versus \emph{Results}. Search results were indexed by a unique Digital Object Identifier (DOI). Attribute data collected per DOI included journal volume and subject classification(s).

@@

The ANZCTR was established in 2005 as part of a coordinated global effort to improve research quality and transparency in clinical trials reporting; observational studies can also be registered. All studies registered on ANZCTR are publicly available and can be searched via an online portal (\url{https://www.anzctr.org.au}).
Details required for registration follow a standardised template (ANZCTR \protect\hyperlink{ref-ANZCTR}{2019}), which covers participant eligibility, the intervention(s) being evaluated, study design and outcomes. The information provided must be in English. Studies are not peer reviewed.

For the statistical methods section, researchers are asked to provide a ``brief description'' of the sample size calculations, statistical methods and planned analyses, although this section is not compulsory (ANZCTR \protect\hyperlink{ref-ANZCTR}{2019}). Studies are reviewed by ANZCTR staff for completeness of key information, which does not include the completeness of the statistical methods sections.

All studies available on ANZCTR were downloaded on 1 February 2020 in XML format.
We used all the text available in the ``Statistical methods'' section. We also collated basic information about the study including the study type (interventional or observational), submission date, number of funders and target sample size. These variables were chosen as we believed they might influence the completeness of the statistical methods section, because we expected larger studies and those with funding to be more complete, and we also were interested in changes over time.
Studies prior to 2013 were excluded as the statistical methods section appeared to be introduced in 2013.
Some studies were first registered on the alternative trial database \emph{clinicaltrials.gov} and then also posted to ANZCTR. We excluded these studies because they almost all had no completed statistical methods section as this section is not included in \emph{clinicaltrials.gov}.

@@

We applied the same text cleaning to both data sources.
Text cleaning aimed to standardise notation and statistical terminology, whilst minimising changes to article style and formatting. \emph{R} code used for data extraction and cleaning is available from \url{https://github.com/agbarnett/stats_section}.

Mathematical notation including Greek letters was converted from Unicode characters to plain text. For example, the Unicode characters corresponding to \(\theta\) (\textless U+03B8\textgreater) were replaced with `theta'. Similarly, common symbols outside of Unicode blocks including `\%' (percent) and `\textless{}' (`less-than') were converted into plain text using the `textclean' package (Rinker \protect\hyperlink{ref-textclean}{2018}). General formatting was removed, this included carriage returns, punctuation marks, in-text references (e.g.~``{[}42{]}'') centred equations, and other non-ASCII characters. Text contained inside brackets was retained to maximise content for analysis, with brackets removed.

We compiled an extensive list of statistical terms to standardise descriptions of statistical methods reported across both datasets. An initial list was compiled by calculating individual word frequencies and identifying relevant terms that appeared at least 100 times. Further terms were sourced from index searches of three statistics textbooks (Dobson and Barnett \protect\hyperlink{ref-Dobson2018}{2018}, @Diggle2013, @Bland2015). The final list is provided as Supplementary Material. Plurals (e.g., `chi-squares') unhyphenated (e.g., `chi square') and combined (e.g.~`chisquare') terms were transformed to singular, hyphenated form (e.g., `chi-square'). Common statistical tests were also hyphenated (e.g., `hosmer lemeshow' to `hosmer-lemeshow').

As a final step, common stop words including pronouns, contractions and selected prepositions were removed. We retained selected stop words that, if excluded, may have changed the context of statistical methods being described, for example `between' and `against'.

@@

\subsection{Clustering algorithm}

Let \(P \in R^{M \times N}\) denotes the paper content matrix where the statistical methods sections of \(N\) papers consist of \(M\) distinct terms. Text clustering for identifying common writing themes in these papers requires this matrix to represent with a vector space model. Let Matrix \(P\) be modeled with the unique terms in the processed sections represented with the tf*idf (term frequent * inverse document frequency) weighting schema to consider both common and rare terms.

Text clustering is known to face the curse of dimensionality due to the high number of terms in doc\(\times\)term matrix representation ({\textbf{???}}). Therefore text-based methods based on distance, density or probability face difficulties ({\textbf{???}}). Specifically, the distance difference between near and far points becomes negligible in high-dimensional data ({\textbf{???}}). This directly affects the distance-based methods such as \textit{k}-means \cite{jain} in accurately identifying the common subgroups. In addition, the sparseness of this high dimensional matrix representation does not allow to differentiate the user groups based on density differences ({\textbf{???}}).

Non-negative Matrix factorization (NMF) which maps the high-dimensional data to a lower-dimensional space has been found to provide an effective solution by allowing to form clusters in the lower-dimensional space ({\textbf{???}}). NMF approximates topic vectors in linear manner considering the context of terms.

In traditional NMF ({\textbf{???}}), the high-dimensional matrix \(P \in R^{M \times N}\) is approximated by learning two factor matrices \(W \in R^{M \times g}\) and \(H \in R ^{N \times g}\) where \(g\) is the number of cluster groups or common topics in the data collection, as follows.

\begin{equation}
\label{eq:2}
 P \approx WH^{T} 
\end{equation}

The matrix factorization process approximates the lower dimensional non-negative factor matrices \(W\) and \(H\) such that they can represent high dimensional \(P\) with the least error. The objective function in Eq \ref{eq:Eq2} iteratively attempts to update the entries in \(W\) and \(H\) to find the optimum values which possess minimum sum of square error for all the elements in both of those matrices. NMF uses Frobenius norm as its objective function and will find the optimum value of \textit{W} and \textit{H} iteratively.

\begin{equation}
\label{eq:Eq2}
\min \frac{1}{2}\|P - WH\|= \sum _{i=1}^{M}\sum _{j=1}^{N} \left(  P_{i,j} -\left(WH \right)_{i,j} \right)^{2}
\end{equation}

Matrix \textit{H} contains the information regarding topic membership of each document. Topic membership of each method section in \(P\) is obtained considering the maximum coefficient value in \(H\) for a method section.

We applied the NMF clustering algorithm to the cleaned dataset, varying the number of clusters from 1 to 50. Clustering solutions were assessed using silhouette score, within-cluster dispersion and between-cluster
dispersion.

For each dataset, records assigned to invididual clusters were examined for evidence of boilerplate text in two ways. We first reviewed the strongest matches to each cluster based on topic coherence. We also compared similarities in text between pairs of records assigned to each cluster. Similarities were determined from calculating cosine similarity scores, with comparisons limited to the top 500 records per cluster. Higher similarity score represent greater overlap in text content between statistical methods sections. Results were transformed to lower case for the clustering, but examples are given using the original capitalisation.

@@

The statistical methods section for the ANZCTR data was missing for some studies and we examined if there were particular studies where this section was more likely to be missing.
We used four independent variables of date, study type (observational or interventional), number of funders and target sample size.
We used a logistic regression model fitted using a Bayesian paradigm. A small number of sections were labelled as ``Not applicable'', ``Nil'' or ``None'' and we changed these to missing.

@@

\hypertarget{plos-one}{%
\subsection{\texorpdfstring{\emph{PLOS ONE}}{PLOS ONE}}\label{plos-one}}

\begin{figure}
\centering
\includegraphics{draft_manuscript_files/figure-latex/unnamed-chunk-3-1.pdf}
\caption{\label{fig:unnamed-chunk-3}\label{fig:plos-n}Search results by \emph{PLOS ONE} volume (1st row); word count per statistical methods section included in analysis (n = 111,731; 2nd row); subject classifications assigned to full-text records included in analysis (3rd row)}
\end{figure}

API searches returned 131,847 unique records, of which 111,731 (85\%) included a statistical methods section based on our search criteria. In the final sample, 95,518 (85\%) DOIs returned an exact match against common section headings, including 64,133 for `statistical analysis', 13,380 for `statistical analyses' and 13,627 for `data analysis'. Among DOIs that did not meet the partial matching criteria, initial search terms appeared in {[}TODO{]}.

Search results varied by journal volume (Figure \ref{fig:plos-n}A). The total number of API search results peaked at volumes 8 (n = 19,045) and 9 (n = 19,045), corresponding to years 2013 and 2014. This trend aligned with the total number of papers published in \emph{PLOS ONE} over the same period. The percentage of records that included a statistical methods section by volume based on our proposed matching criteria varied between 64\% (volume 2) and 86\% (volume 9).

The median length of statistical methods sections was 127 words (IQR: 61 to 253 words) (Figure \ref{fig:plos-n}B). 7,348 articles (7\%) had a statistical methods section of 500 words or more. 19,479 articles (17\%) had a statistical methods section of 50 words or less, equal to the length of this paragraph.

All papers included Biology and life sciences (n = 107,584), Earth sciences
(n = 7,605) and/or Computer and information sciences (n = 5,190) in their top 3 subject classifications
(Figure \ref{fig:plos-n}C).

We applied the clustering algorithm to the cleaned dataset, varying the number of clusters from 1 to 50. Increasing the number of clusters decreased cluster quality based on global goodness-of-fit measures (Supplementary Figure 1), with average silhouette score and within-cluster dispersion leveling off around 20 clusters. This indicated that the data comprised one large, heterogeneous cluster and multiple smaller clusters.

\includegraphics[width=3.75in,height=2.75in,keepaspectratio]{draft_manuscript_files/figure-latex/unnamed-chunk-4-1.png}

The topic clouds based on ten clusters are in Figure\textasciitilde\ref{fig:plot-10-topics}. Frequently occurring words reflected the use of statistical software (Topics\textasciitilde3 and 5), descriptive statistics (Topic\textasciitilde6), group based hypothesis testing (Topics 1 and 4) and definitions of statistical significance (Topics\textasciitilde1 and 9). There are also statistical methods sections associated with regression (Topic\textasciitilde2) and meta-analysis (Topic\textasciitilde7).

Topics related to the use of statistical software differentiated between Prism GraphPad (Topic\textasciitilde3: n = 9,879; 8.8\%) and SPSS (Topic 5: n = 9,574; 8.6\%) (Box 1). A manual review of the top matching sections in these topics showed strong evidence of boilerplate text. Nine out ten top matches for Topic 3 stated the use of Prism GraphPad but did not specify which statistical methods were used; six out of ten top matches returned the same cluster score indicating near identical text. Top matching sections for Topic 5 included information on SPSS version numbers and definitions of statistical significance.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\textbf{Box 1: Examples of boilerplate text from \emph{PLOS ONE} (Statistical software)}

Topic 3

\begin{itemize}
\tightlist
\item
  \emph{graphpad prism (graphpad software, san diego, ca) was used for all analyses.}
\item
  \emph{all statistical analysis was performed using the graphpad prism software.}
\item
  \emph{statistical analysis were done using graphpad prism software (graphpad software, inc., la jolla, ca).}
\end{itemize}

Topic 5

\begin{itemize}
\tightlist
\item
  \emph{all statistical analyses were conducted in spss 19 (spss inc., chicago, il, usa).}
\item
  \emph{all statistical analyses were performed using the spss version 15 for windows (spss inc., chicago, il, usa). a p-value less than 0.05 was considered a significant difference.}
\item
  \emph{statistical analysis was performed using the spss 17.0 software package (spss, chicago, il, usa). all data were analyzed using t tests, values of p \textless{} 0.05 were considered significant.}
\end{itemize}

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

Definitions of statistical significance featured strongly in Topic\textasciitilde1 (n = 3,775) and Topic 9 (n = 6,104) combined with hypothesis testing for comparing differences between two groups. Topic\textasciitilde1 reflected applications of Student's t-test assuming a 5\% level of statistical significance. Topic\textasciitilde9 referenced similar methods combined with multiple thresholds for declaring statistical significance by asterisk: ``*p\textless0.05, **p\textless0.01 and ***p\textless0.001'', a practice that has been criticised (Wasserstein, Schirm, and Lazar \protect\hyperlink{ref-Wasserstein2019}{2019}). Examples of boilerplate text are in Box\textasciitilde2.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\textbf{Box 2: Examples of boilerplate text from \emph{PLOS ONE} (Between-group hypothesis testing, two groups)}

Topic 1

\begin{itemize}
\tightlist
\item
  \emph{student's t-test was used for statistical analysis. a p value of \textless0.05 was considered statistically significant.}
\item
  \emph{statistical analysis was performed using student`s t-test. a p-value \textless0.05 was considered to be statistically significant.}
\item
  \emph{statistical analysis was performed using a two-tailed student's t-test, and p-values \textless0.05 were considered statistically significant.}
\end{itemize}

Topic 9

\begin{itemize}
\tightlist
\item
  \emph{statistical significance was determined by students t-tests. * p\textless0.05, ** p = \textless0.01, *** p\textless0.001.}
\item
  \emph{statistical analyses were performed using the nonparametric mann-whitney test. differences were considered significant for} *\emph{p\textless0.05,} ** \emph{p\textless0.01, *** }p\textless0.001.*
\item
  \emph{comparisons between two groups were performed by student's test. statistical significance was defined as *}p\textless0.05;* ** \emph{p\textless0.01;} *** \emph{p\textless0.001.}
\end{itemize}

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}
